A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-11-07 latex/main.tex
Updates from Overleaf

\author[1,7,*,\orcidlink{0000-0002-6351-8873}]{Shing~H.~Zhan}
\author[1,6,\orcidlink{0000-0002-3536-6411}]{Yan~Wong}
\author[2,6,\orcidlink{0000-0001-6402-9396}]{Anastasia~Ignatieva}
\author[3]{Katherine~Eaton}
\author[1]{Isobel~Guthrie}
\author[1]{Benjamin~Jeffery}
\author[1,2,4,\orcidlink{0000-0002-0824-0047}]{Duncan~S.~Palmer}
\author[3,\orcidlink{0000-0002-1543-4501}]{Carmen~Lia~Murall}
\author[5,\orcidlink{0000-0003-3042-0818}]{Sarah~P.~Otto}
\author[1,**,\orcidlink{0000-0002-7894-5253}]{Jerome~Kelleher}

@@

Millions of SARS-CoV-2 genome sequences were collected 
during the COVID-19 pandemic, forming a dataset of unprecedented richness.
Estimated genealogies are fundamental to
understanding this ocean of data 
and form the primary input to many downstream analyses.
A basic assumption of methods to infer genealogies from viral 
genetic data is that recombination is negligible and 
the genealogy is a tree.
However, recombinant lineages have risen to global prevalence, 
and simple tree representations are therefore 
incomplete and potentially misleading.
We present sc2ts, a method to infer reticulate genealogies 
as an Ancestral Recombination Graph (ARG) in real time at pandemic scale.
We infer an ARG for 2.48 million SARS-CoV-2 genomes 
which leverages the widely used tskit software ecosystem to support further analyses and 
visualisation.
This rich and validated ARG resource clarifies
the relationships among recombinant lineages,
quantifies the rate of recombination over time,
and provides a 
lower bound on detectable recombination.

@@

Ancestral recombination graphs (ARGs) describe the reticulate 
genealogies of sampled sequences in a recombining species\cite{Wong2024},
and provide a natural and efficient means of 
systematically incorporating recombination 
into the study of SARS-CoV-2.
While ARGs have been of theoretical interest in population genetics 
for several decades\cite{Hudson1983,Griffiths1991two,Griffiths1997ancestral},
recent breakthroughs in
simulation\cite{Kelleher2016,Kelleher2018efficient,Haller2019,Gopalan2025bridging}
and inference~\citep{Rasmussen2014,Speidel2019,Kelleher2019,
Wohns2022,Zhang2023,Gunnarsson2024,Deng2025robust} have led to a surge 
of interest in their application in population and statistical 
genetics\cite{Brandt2024,Lewanski2024,Nielsen2025}.
% This is necessary because a bunch of the phylo literature refers
% to ARGs, but it's all in the "coalescent with recombination" sense
% Rannala2025 dismisses ARG-based methods at a stroke because they're
% all based on the assumption of a single panmictic population 
% TODO could phrase this better, Sally didn't like it.
As the range of applications for ARGs has broadened, the term is no
longer synonymous with the coalescent with 
recombination population genetic model \cite{Hudson1983,Griffiths1997ancestral}
and an ARG is now more usually regarded as a \emph{data structure}
for representing recombinant ancestry\cite{Minichiello2006mapping,Wong2024}.
Strong software support for ARGs has facilitated these developments,
and the tskit library~\cite{Ralph2020} is the basis of a 
rapidly growing software ecosystem of
simulation~\cite{Baumdicker2022,Haller2019,Petr2023,Tsambos2023link,Tagami2024tstrait},
visualisation~\cite{Karthikeyan2025tsbrowse,Kitchens2025tskit,Talbot2025argscape},
inference~\cite{Kelleher2019,Wohns2022,Mahmoudi2022,Martin2025model,Pieszko2025detecting}
% More here? I'm sure there's others
and data analysis~\cite{Fan2022,Nowbandegani2023extremely,Fritze2024forest}
tools.

Here we present sc2ts, a new method to infer ARGs for SARS-CoV-2  
that incrementally adds samples over time in a similar manner 
to UShER, but also automatically detects and incorporates recombination
in real-time. Based on the Viridian dataset, a freely available  
resource purged of various systematic errors and artifacts~\cite{Hunt2024},
we infer an ARG with 2.48 million whole genome sequences.
We show that the non-recombinant portions of this ARG are highly congruent with the UShER phylogeny built 
from the same dataset as well as the Pango nomenclature system~\cite{Rambaut2020},
and that the ARG accurately captures evolutionary features such as mutational 
spectra~\cite{Bloom2023} and deletions\cite{Li2023indels}.
We show that sc2ts automatically and accurately detects details about
known recombinant lineages that were painstakingly collated through a community effort, and provides novel insights into the relationships between these
recombinant lineages. We compare the recombination events detected by 
sc2ts with the state-of-the-art combination of UShER and 
RIPPLES~\cite{Turakhia2022}, showing that sc2ts is more sensitive,
detecting more of the well-characterized recombination events 
curated by the Pango lineage designation community.
The output of sc2ts---incorporating
genealogical relationships, point mutations, deletions, and 
recombination---provides a basis for the study of SARS-CoV-2
that systematically accounts for all of these evolutionary forces.
This ARG, coupled with the rich suite of supporting software
in the tskit~\cite{Ralph2020} and VCF-Zarr~\cite{Czech2025} ecosystems,
enables pandemic-scale analysis on a laptop. 
These resources provide a strong platform for future research,
helping us to learn from the vast troves of data collected during the
pandemic, and to prepare for the next.

@@

We compared the performance of sc2ts recombinant detection 
against RIPPLES run on the Viridian UShER phylogeny,
subset to the 2,475,418 samples shared with the ARG (STAR Methods).
To enable exact comparison, we omitted the QC filtering steps from both
methods, and initially set the RIPPLES ``parsimony improvement'' 
parameter to $p=4$ (rather than the more lenient default of 3), 
to match the equivalent $k=4$ setting used in sc2ts (STAR Methods). 
Under this setting,
RIPPLES reported 1174 recombination events of which 
601 were validated by 3SEQ (51\%; STAR Methods).
% JK: There's no point in making a meal of the 3SEQ validation rate,
% This is notably lower than the equivalent 
% concordance rate of 57\% among the 855 recombination events reported by sc2ts.
Aligning the events reported by both methods using the sets of descendant
samples (STAR Methods), 468 of the events are shared by both,
with the intersection comprising 40\% of the RIPPLES events
and 55\% of the sc2ts events (Figure~\ref{fig:ripples_sc2ts_events}). 
The 706 recombination events output by RIPPLES that do not 
correspond to a recombination event in the sc2ts ARG can largely
be explained by a combination of the choice of sites 
used for inference, and the strict time ordering imposed
by sc2ts (Document~\ref{sec:detailed_ripples}).
The 387 events detected by sc2ts and not by UShER+RIPPLES
would require a detailed dissection of the UShER tree
to explain, which is beyond the scope of this work.
However, the set does contain some well characterised events 
associated with the Pango X lineages.
Table~\ref{tab:ripples_x_lineages} shows the events detected 
by UShER+RIPPLES associated with samples assigned to Pango X
lineages, showing that at $p=4$
only 9 of the Pango X lineages are present and
with unequivocal recombinant origin for 6 
(contrast with 16 Type I events in Table~\ref{tab:pango_x_lineages}).
We therefore reran RIPPLES with $p=3$, resulting in 4111 events overall,
and increasing the intersection of events with sc2ts to 595 (70\%;
Figure~\ref{fig:ripples_sc2ts_events}).
At this more lenient threshold, 23 of the 44 Pango X lineages
are present in the dataset, with unequivocal origins for 10 lineages, and
7 lineages associated with two competing events 
(Document~\ref{sec:detailed_ripples}).
XB and XAJ are among the lineages detected by UShER+RIPPLES,
strengthening evidence that these are false negatives 
for sc2ts (Table~\ref{tab:pango_x_lineages}).

@@

We have focused on applying sc2ts to the Viridian dataset~\cite{Hunt2024} rather 
than the larger GISAID resource~\cite{Shu2017} for two reasons.
Firstly, the Viridian dataset is of higher quality and major efforts
have been made to remove sources of error. Applying sc2ts to 
GISAID data is straightforward, but significant additional efforts 
would be required to address data quality issues.
Secondly, the Viridian data is publicly available and redistributable,
enabling us to share the resulting ARG without restrictions.

@@

[FIXME: add all acknowledgments]

>>>>

\author[1,7,*,\orcidlink{0000-0002-6351-8873}]{Shing~H.~Zhan}
\author[1,6,\orcidlink{0000-0002-3536-6411}]{Yan~Wong}
\author[2,6,\orcidlink{0000-0001-6402-9396}]{Anastasia~Ignatieva}
\author[3]{Katherine~Eaton}
\author[1]{Isobel~Guthrie}
\author[1,\orcidlink{0000-0002-1982-6801}]{Benjamin~Jeffery}
\author[1,2,4,\orcidlink{0000-0002-0824-0047}]{Duncan~S.~Palmer}
\author[3,\orcidlink{0000-0002-1543-4501}]{Carmen~Lia~Murall}
\author[5,\orcidlink{0000-0003-3042-0818}]{Sarah~P.~Otto}
\author[1,**,\orcidlink{0000-0002-7894-5253}]{Jerome~Kelleher}

@@

Millions of SARS-CoV-2 genome sequences were collected 
during the COVID-19 pandemic, forming a dataset of unprecedented richness.
Estimated genealogies are fundamental to
understanding this ocean of data 
and form the primary input to many downstream analyses.
A basic assumption of methods to infer genealogies from viral 
genetic data is that recombination is negligible and 
the genealogy is a tree.
However, recombinant lineages have risen to global prevalence, 
and simple tree representations are therefore 
incomplete and potentially misleading.
We present sc2ts, a method to infer reticulate genealogies 
as an Ancestral Recombination Graph (ARG) in real time at pandemic scale.
We infer an ARG for 2.48 million SARS-CoV-2 genomes 
which leverages the widely used tskit software ecosystem to support further analyses and 
visualisation.
This rich and validated resource clarifies
the relationships among recombinant lineages,
quantifies the rate of recombination over time,
and provides a 
lower bound on detectable recombination.

@@

Ancestral recombination graphs (ARGs) describe the reticulate 
genealogies of sampled sequences in a recombining species\cite{Wong2024},
and provide a natural and efficient means of 
systematically incorporating recombination 
into the study of SARS-CoV-2.
While ARGs have been of theoretical interest in population genetics 
for several decades\cite{Hudson1983,Griffiths1991two,Griffiths1997ancestral},
recent breakthroughs in
simulation\cite{Kelleher2016,Kelleher2018efficient,Haller2019,Gopalan2025bridging}
and inference~\citep{Rasmussen2014,Speidel2019,Kelleher2019,
Wohns2022,Zhang2023,Gunnarsson2024,Deng2025robust} have led to a surge 
of interest in their application in population and statistical 
genetics\cite{Brandt2024,Lewanski2024,Nielsen2025}.
% This is necessary because a bunch of the phylo literature refers
% to ARGs, but it's all in the "coalescent with recombination" sense
% Rannala2025 dismisses ARG-based methods at a stroke because they're
% all based on the assumption of a single panmictic population 
As the range of applications for ARGs has broadened, the term is no
longer synonymous with the coalescent with 
recombination population genetic model \cite{Hudson1983,Griffiths1997ancestral}
and an ARG is now more usually regarded as a data structure
for representing recombinant ancestry\cite{Minichiello2006mapping,Wong2024}.
Strong software support for ARGs has facilitated these developments,
and the tskit library~\cite{Ralph2020} is the basis of a 
rapidly growing software ecosystem of
simulation~\cite{Baumdicker2022,Haller2019,Petr2023,Tsambos2023link,Tagami2024tstrait},
visualisation~\cite{Karthikeyan2025tsbrowse,Kitchens2025tskit,Talbot2025argscape},
inference~\cite{Kelleher2019,Wohns2022,Mahmoudi2022,Martin2025model,Pieszko2025detecting}
% More here? I'm sure there's others
and data analysis~\cite{Fan2022,Nowbandegani2023extremely,Fritze2024forest}
tools.

Here we present sc2ts, a new method to infer ARGs for SARS-CoV-2  
that sequentially adds samples over time in a similar manner 
to UShER, but also automatically detects and incorporates recombination
in real-time. Based on the Viridian dataset, a freely available  
resource purged of various systematic errors and artifacts~\cite{Hunt2024},
we infer an ARG with 2.48 million whole genome sequences.
We show that the non-recombinant portions of this ARG are congruent with the UShER phylogeny built 
from the same dataset as well as the Pango nomenclature system~\cite{Rambaut2020},
and that the ARG accurately captures evolutionary features such as mutational 
spectra~\cite{Bloom2023} and deletions\cite{Li2023indels}.
We show that sc2ts automatically and accurately detects details about
known recombinant lineages that were painstakingly collated through a community effort, and provides novel insights into the relationships between these
recombinant lineages. We compare the recombination events detected by 
sc2ts with the state-of-the-art combination of UShER and 
RIPPLES~\cite{Turakhia2022}, showing that sc2ts is more sensitive,
detecting more of the well-characterized recombination events 
curated by the Pango lineage designation community.
The output of sc2ts---incorporating
genealogical relationships, point mutations, deletions, and 
recombination---provides a basis for the study of SARS-CoV-2
that systematically accounts for all of these evolutionary forces.
This ARG, coupled with the rich suite of supporting software
in the tskit~\cite{Ralph2020} and VCF-Zarr~\cite{Czech2025} ecosystems,
enables pandemic-scale analysis on a laptop. 
These resources provide a strong platform for future research,
helping us to learn from the vast troves of data collected during the
pandemic, and to prepare for the next.

@@

We compared the performance of sc2ts recombinant detection 
against RIPPLES run on the Viridian UShER phylogeny,
subset to the 2,475,418 samples shared with the ARG (STAR Methods).
To enable exact comparison, we omitted the QC filtering steps from both
methods, and initially set the RIPPLES ``parsimony improvement'' 
parameter to $p=4$ (rather than the more lenient default of 3), 
to match the equivalent $k=4$ setting used in sc2ts (STAR Methods). 
Under this setting,
RIPPLES reported 1174 recombination events of which 
601 were validated by 3SEQ (51\%; STAR Methods).
% JK: There's no point in making a meal of the 3SEQ validation rate,
% This is notably lower than the equivalent 
% concordance rate of 57\% among the 855 recombination events reported by sc2ts.
We then compared
the specific events reported by both methods by aligning them
using sets of descendant samples (STAR Methods),
giving 468 events shared between RIPPLES and sc2ts (Figure~\ref{fig:ripples_sc2ts_events}).
The 706 recombination events output by RIPPLES that do not 
correspond to a recombination event in the sc2ts ARG can largely
be explained by a combination of the choice of sites 
used for inference, and the strict time ordering imposed
by sc2ts (Document~\ref{sec:detailed_ripples}).
The 387 events detected by sc2ts and not by UShER+RIPPLES
would require a detailed dissection of the UShER tree
to explain, which is beyond the scope of this work.
However, the set does contain some well characterised events 
associated with the Pango X lineages.
Table~\ref{tab:ripples_x_lineages} shows the events detected 
by UShER+RIPPLES associated with samples assigned to Pango X
lineages, showing that at $p=4$
only 9 of the Pango X lineages are present and
with unequivocal recombinant origin for 6 
(contrast with 16 Type I events in Table~\ref{tab:pango_x_lineages}).
We therefore reran RIPPLES with $p=3$, resulting in 4111 events overall,
and increasing the intersection of events with sc2ts to 595 (70\%;
Figure~\ref{fig:ripples_sc2ts_events}).
At this more lenient threshold, 23 of the 44 Pango X lineages
are identified as descendants of RIPPLES recombinant events,
with unequivocal origins for 10 lineages, and
7 lineages associated with two competing events 
(Document~\ref{sec:detailed_ripples}).
XB and XAJ are among the lineages detected by UShER+RIPPLES,
strengthening evidence that these are false negatives 
for sc2ts (Table~\ref{tab:pango_x_lineages}).

@@

We have focused on applying sc2ts to the Viridian dataset~\cite{Hunt2024} rather 
than the larger GISAID resource~\cite{Shu2017} for two reasons.
Firstly, the Viridian dataset is of higher quality and major efforts
have been made to remove sources of error. Applying sc2ts to 
GISAID data is straightforward, but significant additional efforts 
would be required to address data quality issues.
% NOTE Sally suggested that we say that we did this (run on GISAID) and cite 
% the earlier preprint, but after discussion it was concluded that
% this opens a major can of worms and we risk undermining the 
% perceived novelty of this manuscript to guard against a very
% minor technical point. If reviewers ask us to do things with
% GISAID we can by all means explain we've already done it.
Secondly, the Viridian data is publicly available and redistributable,
enabling us to share the resulting ARG without restrictions.

@@
